Automatic Parallelization and Transparent Fault Tolerance

### 2016-07-12 pubs/tfp16-draft/paper.tex
Trims Mini-Haskell appendix.  Minor edits therein.

% because # looks strangely large next to math terms

\newcommand{\hash}{{\scriptsize\#}}
\begin{figure}[h!tb]
\centering
\footnotesize % tiny scriptsize footnotesize small
\begin{tabular}{r r c l l}
Variable         & $f,\ x$             &     &  & Initial lower-case letter \\

@@

Atom               & $a$                 & ::= & $\mathit{lit}\ |\ x\ |\ C$ & Constructors are first class \\% not sure if that's the right terminology 
\\

@@

\subsubsection{Removal of explicit heap objects.}
In STG, all object definitions require explicit classification with
\texttt{FUN}, \texttt{PAP}, \texttt{CON}, \texttt{THUNK}. This classification can be
inferred from context by the compiler so these keywords are not necessary.
Functions can then be defined in Haskell style with the formal parameters following
the function's identifier on the left side of the equality. With these simple
changes the syntax becomes a subset of Haskell.

Heap objects are still only created at the top level or in \texttt{let}
expressions, so the translation is straightforward: \texttt{FUN} objects are
identifiable by the inclusion of these parameters.  A \texttt{PAP} object is
created when a known function is partially applied.  \texttt{CON} objects are
created when constructors are fully applied.  Every other object becomes a
\texttt{THUNK}.

@@

The inclusion of the layout rule may seem insignificant, but it
allows users to produce code that is visually similar to Haskell code.
This makes it easier to read and write for any Haskell programmer who relies
on layout to delineate context, presumably the vast majority.  It also allows
for the use of Haskell programming environments that support ``smart''
indentation, e.g., \emph{haskell-mode} for Emacs~\cite{haskellmode}.

@@

In STG function application must be to atomic values, i.e., either literals or
variables.  This often requires introducing bindings with either \texttt{let} or
\texttt{case} expressions.  Consider the \texttt{append} function that
concatenates its two list parameters, which can be expressed in STG by
{\footnotesize
\begin{verbatim}
   append = FUN(l1 l2 ->
               case l1 of
                 { Nil -> l2;
                   Cons hd tl -> 
                     let { rec = THUNK(append tl l2);
                           result = CON(Cons hd rec) }
                     in result })
\end{verbatim}}
\noindent In mini-Haskell expressions may be applied to other expressions, so the
recursive call to \texttt{append} can be parenthesized. There is no need for the
\texttt{let} expression:
{\footnotesize
\begin{verbatim}
   append l1 l2 = case l1 of
                    Nil -> l2
                    Cons hd tl -> Cons hd (append tl l2)
\end{verbatim}}
\noindent This also serves as a good example of how expression to expression application is
transformed into STG.  Any non-atomic expression in an application expression is
bound to a heap object in a generated \texttt{let} expression.  These
expressions are thus properly ``atomized'' for STG.  The above mini-Haskell code
is transformed into STG that is structurally identical to the STG code that
precedes it.

@@

STG supports pattern matching but only in a simple form and only in
\texttt{case} expressions.  Mini-Haskell extends this to something approaching
what is found in Haskell.  Functions, including lambda expressions, may match on
all their arguments. Named functions may have multiple definitions to exhaust
the combinations of patterns in their arguments and patterns in all contexts
(functions and case expressions) may be nested arbitrarily deep.  This feature
greatly simplifies the task of defining more complex behavior.  Without it,
each matched parameter of a function requires an additional level of nested
\texttt{case} expressions. The same is true of each level of nesting within a
single pattern.
%
Even relatively simple functions can be painful to write, such as \texttt{unzip}:
{\footnotesize
\begin{verbatim}
   unzip = FUN(list ->
     case list of { Nil -> let {res = CON( TP2 nil nil)} in res;
                    Cons x xs ->
                      case x of {
                        TP2 a b ->     --TP2 is the pair constructor
                          case unzip xs of {
                            TP2 as bs ->
                              let {aas = CON(Cons a as);
                                   bbs = CON(Cons b bs);
                                   res = CON(TP2 aas bbs)}
                              in res }}})
\end{verbatim}}
\noindent In mini-Haskell this becomes
{\footnotesize
\begin{verbatim}
   unzip :: List (Tupl2 a b) -> Tupl2 (List a) (List b)
   unzip Nil = TP2 Nil Nil
   unzip (Cons (TP2 a b) xs) =
     case unzip xs of
       TP2 as bs -> TP2 (Cons a as) (Cons b bs)
\end{verbatim}}

Of course, this complex pattern matching must eventually be transformed into
the primitive, nested case expressions. Fortunately there are well-documented
algorithms for achieving this, and Philip Wadler's chapter
% in \emph{The Implementation of Functional Programming Languages}
on the compilation of this type of pattern matching proved
invaluable~\cite{Jones:1987}.  Wadler describes the \emph{match} function
which, in its simplest form, produces a correct, but potentially large
translation with some redundancy.  This function was implemented with only
minor modifications to fit the shape of mini-Haskell's abstract form.  The
implementation is augmented by some light flow-analysis to eliminate some of
the redundancy and make the resulting code easier to read.  There are further
optimizations suggested by Wadler to address the size of the generated code,
for example, when multiple branches lead to the same large expression, but the
additional complexity outweighs the benefit for a testing language.

@@

Mini-Haskell provides some smaller conveniences that don't warrant individual
sections but are worth mentioning.
\begin{itemize}
\item \emph{Constructors as functions.}
  As in Haskell, constructors are first class, acting as true functions.  This
  is implemented by simply generating a function that wraps the appropriate
  \texttt{let}-bound \texttt{CON} object in STG.  Any partially applied
  constructor generates a call to this function.  Fully applied constructors
  generate local \texttt{let}-bound objects to avoid an unnecessary function
  call.

\item \emph{Primitive operations as functions.}
  Similar to constructors, primitive operations are given first-class status as
  functions, with roughly the same rules.  Partial applications are
  translated to calls to generated functions that wrap the actual primitive
  operations. Fully applied calls generate true primitive operation calls.

\item \emph{Lambda expressions.}
  mini-Haskell supports anonymous lambda expressions for the simple, one-off
  functions that often find a place in Haskell code.  This was certainly the
  simplest feature to add, requiring only that the function be given a unique
  identifier to bind it to a \texttt{FUN} object in STG.

@@

% I don't think I've described this very well at all... - david
\item \emph{Haskell Block.}
  Being a subset of Haskell, it may be desirable to test a mini-Haskell program
  compiled through our toolchain against the same program (and same source file)
  compiled with another Haskell compiler.  Because mini-Haskell exposes a
  different set of primitives the Haskell Block was introduced to allow the
  inclusion of the \texttt{MagicHash} extension and the aliasing Haskell's
  primitives (e.g. \texttt{(+)} and \texttt{(-)}) to the names given in
  mini-Haskell.  This block, delimited by \texttt{\{-\#-\}} strings, is ignored
  by the mini-Haskell compiler, and arbitrary ``real'' Haskell code can be
  placed therein to create the same mini-Haskell environment for full Haskell
  compilers.

>>>>

% because # looks strangely large next to math terms
\newcommand{\hash}{{\scriptsize\#}}

\begin{figure}[h!tb]
\centering
\footnotesize % tiny scriptsize footnotesize small
\begin{tabular}{r r c l l}
Variable         & $f,\ x$             &     &  & Initial lower-case letter \\

@@

Atom               & $a$                 & ::= & $\mathit{lit}\ |\ x\ |\ C$ & Constructors are first class \\
\\

@@

\subsubsection{Removal of explicit heap objects.}
In STG, all object definitions require explicit classification with
\texttt{FUN}, \texttt{PAP}, \texttt{CON}, \texttt{THUNK}. This
classification can be inferred from context by the compiler so these
keywords are not necessary.  Functions can then be defined in Haskell
style with the formal parameters following the function's identifier
on the left side of the equality. With these simple changes the syntax
becomes a subset of Haskell. Heap objects are still only created at
the top level or in \texttt{let} expressions, so the translation is
straightforward.

\begin{comment} Trimming Mini-Haskell to fit page limit
\texttt{FUN} objects are
identifiable by the inclusion of these parameters.  A \texttt{PAP} object is
created when a known function is partially applied.  \texttt{CON} objects are
created when constructors are fully applied.  Every other object becomes a
\texttt{THUNK}.
\end{comment}

@@

\begin{comment} Trimming Mini-Haskell to fit page limit
The inclusion of the layout rule may seem insignificant, but it
allows users to produce code that is visually similar to Haskell code.
This makes it easier to read and write for any Haskell programmer who relies
on layout to delineate context, presumably the vast majority.  It also allows
for the use of Haskell programming environments that support ``smart''
indentation, e.g., \emph{haskell-mode} for Emacs~\cite{haskellmode}.
\end{comment}

@@

In STG function application must be to atomic values, i.e., either literals or
variables.  This often requires introducing bindings with either \texttt{let} or
\texttt{case} expressions.  Consider the \texttt{append} function that
concatenates its two list parameters, which can be expressed in STG by
{\footnotesize
\begin{verbatim}
   append = FUN(l1 l2 -> case l1 of x {
               Nil -> l2; 
               Cons hd tl -> let { rec = THUNK(append tl l2);
                                   result = CON(Cons hd rec) }
                             in result })
\end{verbatim}}
\noindent In mini-Haskell expressions may be applied to other expressions, so the
recursive call to \texttt{append} can be parenthesized. There is no need for the
\texttt{let} expression:
{\footnotesize
\begin{verbatim}
   append l1 l2 = case l1 of
                    Nil -> l2
                    Cons hd tl -> Cons hd (append tl l2)
\end{verbatim}}
\noindent This also serves as a good example of how expression to expression application is
transformed into STG.  Any non-atomic expression in an application expression is
bound to a heap object in a generated \texttt{let} expression.  These
expressions are thus properly ``atomized'' for STG.  The above mini-Haskell code
is transformed into STG that is structurally identical to the STG code that
precedes it.

@@

STG supports pattern matching but only in a simple form and only in
\texttt{case} expressions.  Mini-Haskell extends this to something approaching
what is found in Haskell.  Functions, including lambda expressions, may match on
all their arguments. Named functions may have multiple definitions to exhaust
the combinations of patterns in their arguments, and patterns in all contexts
(functions and case expressions) may be nested arbitrarily deep.  This feature
greatly simplifies the task of defining more complex behavior.  Without it,
each matched parameter of a function requires an additional level of nested
\texttt{case} expressions. The same is true of each level of nesting within a
single pattern.
%
Even relatively simple functions can be painful to write, such as \texttt{unzip}:
{\footnotesize
\begin{verbatim}
   unzip = FUN(list -> 
     case list of s1 { Nil -> let {res = CON( Pair nil nil)} in res;
                    Cons x xs -> case x of s2 {
                      Pair a b -> case unzip xs of s3 {
                        Pair as bs -> let {aas = CON(Cons a as);
                                           bbs = CON(Cons b bs);
                                           res = CON(Pair aas bbs)}
                                      in res }}})
\end{verbatim}}
\noindent In mini-Haskell this becomes
{\footnotesize
\begin{verbatim}
   unzip :: List (Pair a b) -> Pair (List a) (List b)
   unzip Nil = Pair Nil Nil
   unzip (Cons (Pair a b) xs) =
     case unzip xs of
       Pair as bs -> Pair (Cons a as) (Cons b bs)
\end{verbatim}}

Of course, this complex pattern matching must eventually be transformed into
the primitive, nested case expressions. Fortunately there are well-documented
algorithms for achieving this, and Philip Wadler's chapter
% in \emph{The Implementation of Functional Programming Languages}
on the compilation of this type of pattern matching proved
invaluable~\cite{Jones:1987}.

\begin{comment} Trimming Mini-Haskell to fit page limit
Wadler describes the \emph{match} function
which, in its simplest form, produces a correct, but potentially large
translation with some redundancy.  This function was implemented with only
minor modifications to fit the shape of mini-Haskell's abstract form.  The
implementation is augmented by some light flow-analysis to eliminate some of
the redundancy and make the resulting code easier to read.  There are further
optimizations suggested by Wadler to address the size of the generated code,
for example, when multiple branches lead to the same large expression, but the
additional complexity outweighs the benefit for a testing language.
\end{comment}

@@

Mini-Haskell provides some smaller conveniences that don't warrant individual
sections but are worth mentioning.
\begin{itemize}
\item \emph{Constructors as functions.}
  As in Haskell, constructors are first class, acting as true functions.  This
  is implemented by simply generating a function that wraps the appropriate
  \texttt{let}-bound \texttt{CON} object in STG.  %% Any partially applied
  %% constructor generates a call to this function.  Fully applied constructors
  %% generate local \texttt{let}-bound objects to avoid an unnecessary function
  %% call.

\item \emph{Primitive operations as functions.}
  Similar to constructors, primitive operations are given first-class status as
  functions, with roughly the same rules.  %% Partial applications are
  %% translated to calls to generated functions that wrap the actual primitive
  %% operations. Fully applied calls generate true primitive operation calls.

\item \emph{Lambda expressions.}
  mini-Haskell supports anonymous lambda expressions for the simple, one-off
  functions that often find a place in Haskell code.%%   This was certainly the
  %% simplest feature to add, requiring only that the function be given a unique
  %% identifier to bind it to a \texttt{FUN} object in STG.

@@

% I don't think I've described this very well at all... - david
\item \emph{Haskell Block.}
  Being a subset of Haskell, it may be desirable to test a mini-Haskell program
  compiled through our toolchain against the same program (and same source file)
  compiled with another Haskell compiler.  Because mini-Haskell exposes a
  different set of primitives the Haskell Block was introduced to allow the
  inclusion of the \texttt{MagicHash} extension and the aliasing Haskell's
  primitives (e.g. \texttt{(+#)} and \texttt{(-#)}) to the names given in
  mini-Haskell.  This block, delimited by \texttt{\{-\#-\}} strings, is ignored
  by the mini-Haskell compiler, and arbitrary ``real'' Haskell code can be
  placed therein to create the same mini-Haskell environment for full Haskell
  compilers.
